---
title: 'PHIL 296: Honors Logic (Revised Timetable)'
author: "Brian Weatherson"
date: "Fall 2022"
format:
  pdf:
    documentclass: article
    md-extensions: +compact_definition_lists
    keep-tex: true
    block-headings: false
    pdf-engine: xelatex
    include-in-header: 
      - packages.tex
  html:
    minimal: true
geometry: margin=1.5in
mainfont: SF Pro Thin
mainfontoptions: 
 - BoldFont = SF Pro Rounded Semibold
 - Scale = MatchLowercase
mathfont: Fira Math
editor: source
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='_cache/')
knitr::opts_chunk$set(warning=F, message=F)
# knitr::opts_knit$get(kable.force.latex = TRUE)
require(tidyverse)
require(fontawesome)
```

```{r dates_bib, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, dev='pdf')
mon <- as.Date("2022-08-29") # Monday of the First Week
tue <- mon + 1
wed <- tue + 1
thu <- wed + 1
fri <- thu + 1
# fri <- as.Date("2016-01-08")

advdate <- function(obj, adv) {
 theday <- obj + 7*(adv-1)
 ttheday <- format(theday, format="%A, %B %d")
 return(ttheday)
}

#library(RefManageR)
## library(knitcitations)
## library(rcrossref)
#bib <- ReadBib("411bib.bib")
#myopts <- BibOptions(bib.style = "authoryear", style="latex", first.inits=FALSE, max.names = 20)

```

| **Lead Instructor**: Brian Weatherson
| `r fa("far fa-envelope")` weath\@umich.edu
| `r fa("fas fa-wifi")` canvas.umich.edu
| 
| **Office Hours**: TBA 
| 

# Course Description

This course is a fairly fast-paced introduction to many different logics. We will start with a very quick discussion of **classical logic**, the logic that is used in most other courses (both in philosophy and elsewhere). We'll then move on to **modal logic**, the logic of possibility and necessity. And finally, we'll look at rivals to classical logic, the imaginatively named **non-classical logics**. The aim of the course is to both give you a working familiarity with the technical aspects of these logics, but also some background about why one might care about rivals to classical logic, and when one might think these rivals are suitable to use.

# Canvas

There is a Canvas site for this course, which can be accessed from <https://canvas.umich.edu>. Course documents (syllabus, lecture notes, assignments) will be available from this site. Please make sure that you can access this site. Consult the site regularly for announcements, including changes to the course schedule. And there are many tools on the site to communicate with each other, and with me.

\newpage

# Required Materials

The bulk of the course will be spent working through selections from

-   *An Introduction to Non-Classical Logic: From If to Is*, by Graham Priest Cambridge University Press, 2008. Second Edition.

It is not important that you have the physical version of the book; the electronic version is fine. But it is important that you have the second edition; the first edition only has half the material.

# Course Requirements

1.  Participate in class. We will spend a bit of time in class going over problems, and it is important to practice these.
2.  Do weekly online quizzes. These will be available through Canvas.
3.  Do a final exam. This will also be through Canvas.

# Grade Breakdown

-   Class participation: 10%;
-   Ten Quizzes, of which the best 8 will count: 7.5% each, for a total of 60%;
-   Final Exam: 30%.

The quizzes will be due on Friday, and will mostly cover the material from that week. (The exceptions are noted in the text below.) The exam will be held through Canvas, and will be due on the scheduled date of the exam. The schedule is at <https://ro.umich.edu/calendars/final-exams/2022-fall>, and I believe that means we're to be done by 6pm on Monday, December 12.

\newpage

This is the revised timetable for the second half of the term; after the first one was considerably too ambitious. I hope we can stick to this one, though it again we'll see if we have to adjust


## `r advdate(wed, 8)`

Topic
:   Reading a model off an open tableau in modal logic

Required Reading
:   Priest, §§2.4.7-2.4.9

Weekly Quiz
:   Due 5pm, `r advdate(fri, 8)`

## `r advdate(mon, 9)`

Topic
:   Three more modal logics (plus combinations)

Required Reading
:   Priest, §3.2

## `r advdate(wed, 9)`

Topic
:   Tableau for these new logics

Required Reading
:   Priest, §3.3-4

Weekly Quiz
:   Due 5pm, `r advdate(fri, 9)`


## `r advdate(mon, 10)`

Topic
:   The special logic S5

Required Reading
:   Priest, §3.5

## `r advdate(wed, 10)`

Topic
:   Applications

Required Reading
:   Not in the book!

Weekly Quiz
:   Due 5pm, `r advdate(fri, 10)`

## `r advdate(mon, 11)`

Topic
:   Strict Conditionals

Required Reading
:   Priest, §4.5-4.9

## `r advdate(wed, 11)`

Topic
:   Problems for Strict Conditionals

Required Reading
:   Priest, §5.1-5.2

Weekly Quiz
:   Due 5pm, `r advdate(fri, 11)`

## `r advdate(mon, 12)`

Topic
:   Conditional Semantics

Required Reading
:   Priest, §5.3


## `r advdate(wed, 12)`

Topic
:   Tableau for conditionals

Required Reading
:   Priest, §5.4

Weekly Quiz
:   Due 5pm, `r advdate(fri, 12)`

## `r advdate(mon, 13)`

Topic
:   Similarity Spheres

Required Reading
:   Priest, §5.5-5.6

Weekly Quiz
:   None this week

## `r advdate(mon, 14)`

Topic
:   C1 and C2

Required Reading
:   Priest, §5.7-5.8

## `r advdate(wed, 14)`

Topic
:    Finishing up chapter 5

Required Reading
:    Priest, Ch. 5

Weekly Quiz
:   Due 5pm, `r advdate(fri, 14)`


## `r advdate(mon, 15)`

Topic
:   Review

Required Reading
:   None

- The last week will be for review, and for extra time for any topic we felt was too rushed through the course.

# Plagiarism

Although team-work is encouraged, plagiarism is strictly prohibited. You are responsible for making sure that none of your work is plagiarized. Be sure to cite work that you use, both direct quotations and paraphrased ideas. Any citation method that is tolerably clear is permitted, but if you'd like a good description of a citation scheme that works well in philosophy, look at <http://bit.ly/VDhRJ4>.

You are encouraged to discuss the course material, including assignments, with your classmates, but all written work that you hand in under your own name must be your own. 

You should also be familiar with the academic integrity policies of the College of Literature, Science & the Arts at the University of Michigan, which are available here: <http://www.lsa.umich.edu/academicintegrity/>. Violations of these policies will be reported to the Office of the Assistant Dean for Student Academic Affairs, and sanctioned with a course grade of F.

# Disability

The University of Michigan abides by the Americans with Disabilities Act of 1990, Section 504 of the Rehabilitation Act of 1973, and other applicable federal and state laws that prohibit discrimination on the basis of disability, which mandate that reasonable accommodations be provided for qualified students with disabilities.

If you have a disability, and may require some type of instructional and/or examination accommodation, please contact me early in the semester. If you have not already done so, you will also need to register with the Office of Services for Students with Disabilities. The office is located at G664 Haven Hall.

For more information on disability services at the University of Michigan, go to <http://ssd.umich.edu>.


